---
title: "Bayesian Hierarchical Modeling of NBA Shot-Making Probability"
author: "Rex Nelson, William Ken Li, Andrew Cho"
format:
  html:
    toc: true
    toc-depth: 3
    theme: darkly
    code-fold: true
execute:
  warning: false
  message: false
  echo: true
  code-fold: true
---

# Research question

In basketball, a shot attempt is not equally difficult for every player or every situation. A layup, a short midrange jumper, and a three-point attempt have very different expected outcomes. At the same time, players differ in their individual shooting ability, and players with only a small number of attempts are difficult to evaluate from raw field goal percentage alone.

Our research question is:

> How do shot context and player-specific ability affect the probability that an NBA player makes a field goal attempt?

We analyze shot-level NBA data using a Bayesian hierarchical logistic regression model. The response variable is whether a shot was made. The main predictors are shot distance, shot type, a derived shot zone, period, time remaining in the period, score margin, home/away status, and player identity. The hierarchical part of the model allows each player to have their own shooting ability parameter while partially pooling players toward the league average. This is useful because players with fewer attempts should not be treated as if their observed field goal percentage is as reliable as that of high-volume shooters.

The main goals are:

1.  Estimate how shot context changes the probability of making a shot.
2.  Estimate player-specific shooting ability after adjusting for shot difficulty.
3.  Demonstrate Bayesian partial pooling.
4.  Compare the Bayesian model to a comparable frequentist logistic regression.
5.  Evaluate model fit using posterior prediction, posterior predictive checking, prior sensitivity, and model comparison.

# Statistical background

For each shot attempt (i = 1, \ldots, n), define

$$Y_i =
\begin{cases}
1, & \text{if shot } i \text{ was made} \\
0, & \text{if shot } i \text{ was missed.}
\end{cases}$$

We model

$$Y_i \mid p_i \sim \text{Bernoulli}(p_i),$$

where (p_i) is the probability that shot (i) is made.

Because (p_i) must lie between 0 and 1, we use a logistic regression model:

$$\text{logit}(p_i) = \log\left(\frac{p_i}{1-p_i}\right) = \eta_i.$$

A non-hierarchical model would use

$$\eta_i =
\alpha
+
\beta_1 \,\text{distance}_{z,i}
+
\beta_2 \, I(\text{3PT}_i)
+
\beta_3 \, I(\text{home}_i)
+
\beta_4 \,\text{period}_i
+
\beta_5 \,\text{zone}_i
+
\beta_6 \,\text{clock}_{z,i}
+
\beta_7 \,\text{margin}_{z,i}.$$

However, this ignores player-to-player variation. We instead use a hierarchical model:

$$\eta_i =
\alpha
+
u_{\text{player}[i]}
+
\beta_1 \,\text{distance}_{z,i}
+
\beta_2 \, I(\text{3PT}_i)
+
\beta_3 \, I(\text{home}_i)
+
\beta_4 \,\text{period}_i
+
\beta_5 \,\text{zone}_i
+
\beta_6 \,\text{clock}_{z,i}
+
\beta_7 \,\text{margin}_{z,i}.$$

The player effects are modeled as

$$u_j \sim \text{Normal}(0, \sigma_{\text{player}}),$$

where (u_j) is the adjusted shooting ability of player (j), and (\sigma\_{\text{player}}) describes how much true shooting ability varies across players after controlling for shot context.

This model is Bayesian because we specify priors for all unknown parameters and use posterior inference:

$$p(\theta \mid y) \propto p(y \mid \theta)p(\theta),$$

where (p(y \mid \theta)) is the likelihood and (p(\theta)) is the prior distribution.

# Why this is an appropriate Bayesian project

This project goes beyond a standard logistic regression because it uses a hierarchical Bayesian model. The Bayesian framework is especially useful here for three reasons.

First, the model estimates uncertainty directly using posterior distributions, not just point estimates and standard errors.

Second, partial pooling helps stabilize player estimates. A player who makes 6 out of 10 shots should not necessarily be estimated as a better shooter than a player who makes 420 out of 900 shots. The hierarchical prior shrinks low-information players toward the league average.

Third, Bayesian posterior prediction allows us to answer practical questions such as:

- What is the predicted probability that a specific player makes a three-point attempt?
- How much does shot distance reduce make probability?
- Which players appear to outperform expected shot difficulty?
- Does the model reproduce realistic made-shot rates by zone or player?

# Packages

```{r}
library(tidyverse)
library(hoopR)
library(brms)
library(bayesplot)
library(posterior)
library(lme4)
library(broom.mixed)
library(loo)
library(gt)
library(janitor)
library(forcats)
library(purrr)
```

```{r}
theme_set(theme_minimal(base_size = 13))
options(mc.cores = parallel::detectCores())
set.seed(415)
```

# Data source

We use NBA play-by-play data accessed through the `hoopR` package. The function `load_nba_pbp()` provides play-level data from ESPN, including shot outcomes, shot coordinates, period, clock, score, and player IDs. We also use `load_nba_player_box()` to recover player display names from player IDs when that lookup is available.

This project proposal originally described using Basketball Reference, but the `hoopR` play-by-play data is a better match for a hierarchical shot-level model because it provides consistent shot-level coordinates and game context.

# Data collection

The code below pulls NBA play-by-play data and filters it down to field goal attempts. The ESPN play-by-play file in `hoopR` does not contain a stable `athlete_display_name_1` column, so player names are joined separately using `load_nba_player_box()`. If that lookup fails, the code falls back to labels based on player ID.

The play-by-play file also does not contain a direct `shot_distance` column in your session, so we derive an approximate shot distance from the play text when possible, and otherwise use shot coordinates as a fallback.

```{r}
season_used <- 2024

pbp_raw <- load_nba_pbp(seasons = season_used) |>
  clean_names()

player_lookup <- tryCatch(
  load_nba_player_box(seasons = season_used) |>
    clean_names() |>
    distinct(athlete_id, athlete_display_name),
  error = function(e) {
    tibble(
      athlete_id = integer(),
      athlete_display_name = character()
    )
  }
)

shots_raw <- pbp_raw |>
  mutate(
    shot_text = str_to_lower(coalesce(text, "")),
    parsed_distance = str_extract(shot_text, "\\d+(?=-foot)") |> as.numeric(),
    coord_distance = sqrt(coordinate_x^2 + coordinate_y^2),
    shot_distance = case_when(
      !is.na(parsed_distance) ~ parsed_distance,
      str_detect(shot_text, "dunk") ~ 1,
      str_detect(shot_text, "layup|tip shot|tip-in") ~ 2,
      str_detect(shot_text, "hook shot|floating") ~ 6,
      TRUE ~ coord_distance
    )
  ) |>
  filter(
    shooting_play == TRUE,
    !str_detect(shot_text, "free throw"),
    !is.na(athlete_id_1),
    !is.na(team_id),
    !is.na(home_team_id),
    !is.na(away_team_id),
    !is.na(coordinate_x),
    !is.na(coordinate_y),
    !is.na(period_number),
    !is.na(clock_minutes),
    !is.na(clock_seconds),
    !is.na(home_score),
    !is.na(away_score),
    !is.na(shot_distance)
  ) |>
  transmute(
    made = as.integer(scoring_play == TRUE),
    player_id = athlete_id_1,
    team_id = team_id,
    period = period_number,
    is_home = case_when(
      team_id == home_team_id ~ "home",
      team_id == away_team_id ~ "away",
      TRUE ~ NA_character_
    ),
    shot_distance = shot_distance,
    coordinate_x = coordinate_x,
    coordinate_y = coordinate_y,
    shot_type_simple = case_when(
      score_value == 3 ~ "3PT",
      score_value == 2 ~ "2PT",
      str_detect(shot_text, "three point|3-pt|3pt") ~ "3PT",
      TRUE ~ "2PT"
    ),
    shot_zone_basic = case_when(
      shot_distance <= 4 ~ "Restricted Area",
      shot_distance <= 14 ~ "Paint / Short Midrange",
      shot_distance <= 22 ~ "Long Midrange",
      TRUE ~ "Three Point"
    ),
    score_margin = case_when(
      team_id == home_team_id ~ home_score - away_score,
      team_id == away_team_id ~ away_score - home_score,
      TRUE ~ NA_real_
    ),
    clock_seconds_remaining = clock_minutes * 60 + clock_seconds,
    game_date = game_date
  ) |>
  filter(!is.na(is_home), !is.na(score_margin)) |>
  left_join(player_lookup, by = c("player_id" = "athlete_id")) |>
  mutate(
    player_name = coalesce(athlete_display_name, paste("Player", player_id))
  ) |>
  select(
    made,
    player_id,
    player_name,
    period,
    is_home,
    shot_distance,
    coordinate_x,
    coordinate_y,
    shot_type_simple,
    shot_zone_basic,
    score_margin,
    clock_seconds_remaining,
    game_date
  )
```

```{r}
glimpse(shots_raw)
```

# Restricting the data set

To keep the project computationally reasonable and statistically interpretable, we restrict to players with at least a moderate number of attempts and then sample a few thousand shots. This still leaves enough data for a meaningful hierarchical model while avoiding a huge data set where posterior intervals become extremely small.

```{r}
player_counts <- shots_raw |>
  count(player_name, name = "attempts") |>
  arrange(desc(attempts))

players_keep <- player_counts |>
  filter(attempts >= 60) |>
  slice_head(n = 50) |>
  pull(player_name)

shots_filtered <- shots_raw |>
  filter(player_name %in% players_keep)

shots_model <- shots_filtered |>
  slice_sample(n = min(4000, nrow(shots_filtered))) |>
  mutate(
    distance_z = as.numeric(scale(shot_distance)),
    score_margin_z = as.numeric(scale(score_margin)),
    clock_seconds_z = as.numeric(scale(clock_seconds_remaining)),
    player_name = factor(player_name),
    shot_type_simple = factor(shot_type_simple, levels = c("2PT", "3PT")),
    shot_zone_basic = factor(
      shot_zone_basic,
      levels = c(
        "Restricted Area",
        "Paint / Short Midrange",
        "Long Midrange",
        "Three Point"
      )
    ),
    is_home = factor(is_home, levels = c("away", "home")),
    period = factor(period)
  ) |>
  droplevels()

glimpse(shots_model)
```

# Exploratory data analysis

## Overall make percentage

```{r}
shots_model |>
  summarize(
    n_shots = n(),
    made_shots = sum(made),8
    fg_pct = mean(made)
  ) |>
  gt()
```

## Make percentage by shot type

```{r}
shots_model |>
  group_by(shot_type_simple) |>
  summarize(
    n = n(),
    fg_pct = mean(made),
    .groups = "drop"
  ) |>
  gt()
```

## Make percentage by shot zone

```{r}
shots_model |>
  group_by(shot_zone_basic) |>
  summarize(
    n = n(),
    fg_pct = mean(made),
    .groups = "drop"
  ) |>
  arrange(desc(fg_pct)) |>
  gt()
```

## Shot distance and outcome

```{r}
ggplot(shots_model, aes(x = shot_distance, y = made)) +
  geom_jitter(height = 0.05, alpha = 0.08) +
  geom_smooth(
    method = "glm",
    method.args = list(family = binomial),
    se = TRUE
  ) +
  labs(
    title = "Shot outcome by derived shot distance",
    x = "Shot distance",
    y = "Made shot indicator"
  )
```

## Raw field goal percentage by player

```{r}
player_raw <- shots_model |>
  group_by(player_name) |>
  summarize(
    attempts = n(),
    makes = sum(made),
    raw_fg_pct = mean(made),
    .groups = "drop"
  ) |>
  arrange(desc(raw_fg_pct))

player_raw |>
  slice_head(n = 15) |>
  gt()
```

# Bayesian model

## Likelihood

For shot (i),

$$Y_i \sim \text{Bernoulli}(p_i).$$

## Linear predictor

Our main model is

$$\text{logit}(p_i)
=
\alpha
+
u_{\text{player}[i]}
+
\beta_1 \,\text{distance}_{z,i}
+
\beta_2 I(\text{3PT}_i)
+
\beta_3 I(\text{home}_i)
+
\beta_4 \,\text{period}_i
+
\beta_5 \,\text{clock}_{z,i}
+
\beta_6 \,\text{margin}_{z,i}.$$

The player random effect is

$$u_j \sim \text{Normal}(0, \sigma_{\text{player}}).$$

The interpretation of (u_j) is the player's adjusted shooting ability after controlling for shot distance, shot type, period, time remaining, score margin, and home/away status.

## Priors

We use weakly informative priors.

The intercept receives

$$\alpha \sim \text{Normal}(0, 1.5).$$

On the probability scale, this prior allows a wide range of baseline shooting probabilities. Since NBA field goal percentages are often around 40% to 50%, this is weak but still reasonable.

Regression coefficients receive

$$\beta_k \sim \text{Normal}(0, 1).$$

A coefficient of 1 on the log-odds scale corresponds to an odds ratio of (e\^1 \approx 2.72), so this prior allows substantial effects while discouraging implausibly extreme ones.

The player standard deviation receives

$$\sigma_{\text{player}} \sim \text{Exponential}(1).$$

This prior favors smaller between-player variation but still permits substantial heterogeneity if supported by the data.

```{r}
model_formula <- bf(
  made ~ distance_z + shot_type_simple + is_home + period +
    clock_seconds_z + score_margin_z + (1 | player_name)
)

model_formula_fixed <- bf(
  made ~ distance_z + shot_type_simple + is_home + period +
    clock_seconds_z + score_margin_z
)

priors_main <- c(
  prior(normal(0, 1.5), class = "Intercept"),
  prior(normal(0, 1), class = "b"),
  prior(exponential(1), class = "sd")
)

priors_fixed_only <- c(
  prior(normal(0, 1.5), class = "Intercept"),
  prior(normal(0, 1), class = "b")
)

priors_main
```

# Prior predictive checking

Before fitting the model to the observed outcomes, we check whether the priors imply reasonable shot-making rates.

```{r}
fit_prior <- brm(
  formula = model_formula,
  data = shots_model,
  family = bernoulli(link = "logit"),
  prior = priors_main,
  sample_prior = "only",
  chains = 4,
  iter = 1000,
  seed = 415,
  control = list(adapt_delta = 0.95)
)
```

```{r}
pp_check(fit_prior, ndraws = 100) +
  labs(title = "Prior predictive check")
```

# Fit the main Bayesian hierarchical logistic model

```{r}
fit_bayes <- brm(
  formula = model_formula,
  data = shots_model,
  family = bernoulli(link = "logit"),
  prior = priors_main,
  sample_prior = "yes",
  chains = 4,
  iter = 2500,
  warmup = 1000,
  seed = 415,
  control = list(adapt_delta = 0.97, max_treedepth = 12)
)
```

```{r}
summary(fit_bayes)
```

# MCMC diagnostics

We check standard MCMC diagnostics, including trace plots, (\hat{R}), and effective sample size.

```{r}
plot(fit_bayes)
```

```{r}
draws_array <- as.array(fit_bayes)

mcmc_trace(
  draws_array,
  pars = c(
    "b_Intercept",
    "b_distance_z",
    "b_clock_seconds_z",
    "b_score_margin_z"
  )
)
```

```{r}
draws_df <- as_draws_df(fit_bayes)

diag_summary <- posterior::summarise_draws(
  draws_df,
  posterior::rhat,
  posterior::ess_bulk,
  posterior::ess_tail
) |>
  rename(parameter = variable)

rhat_col <- names(diag_summary)[str_detect(names(diag_summary), regex("rhat", ignore_case = TRUE))][1]
ess_bulk_col <- names(diag_summary)[str_detect(names(diag_summary), regex("ess_bulk", ignore_case = TRUE))][1]
ess_tail_col <- names(diag_summary)[str_detect(names(diag_summary), regex("ess_tail", ignore_case = TRUE))][1]

diag_summary <- diag_summary |>
  rename(
    rhat_value = all_of(rhat_col),
    ess_bulk_value = all_of(ess_bulk_col),
    ess_tail_value = all_of(ess_tail_col)
  ) |>
  filter(str_detect(parameter, "^(b_|sd_|r_)")) |>
  arrange(desc(rhat_value))

diag_summary |>
  slice_head(n = 20) |>
  gt()

```

A well-fit model should have (\hat{R}) values close to 1.00 and no obvious trace plot pathologies. If there are divergent transitions, we would increase `adapt_delta`, simplify the model, or reconsider the parameterization.

# Posterior inference for fixed effects

```{r}
fixef(fit_bayes) |>
  as.data.frame() |>
  rownames_to_column("parameter") |>
  gt()
```

The distance coefficient is expected to be negative because longer shots are generally harder. The home effect, period effects, and game-state effects may be smaller after adjusting for shot distance and player ability.

To interpret coefficients as odds ratios:

```{r}
posterior_fixed <- fixef(fit_bayes) |>
  as.data.frame() |>
  rownames_to_column("parameter") |>
  mutate(
    odds_ratio_estimate = exp(Estimate),
    odds_ratio_q2.5 = exp(Q2.5),
    odds_ratio_q97.5 = exp(Q97.5)
  )

posterior_fixed |>
  select(parameter, odds_ratio_estimate, odds_ratio_q2.5, odds_ratio_q97.5) |>
  gt()
```

# Player-specific posterior estimates

The player effects estimate shooting ability after adjusting for shot context. These are not raw field goal percentages. They are adjusted player effects on the log-odds scale.

```{r}
player_effects <- ranef(fit_bayes)$player_name[, , "Intercept"] |>
  as.data.frame() |>
  rownames_to_column("player_name") |>
  rename(
    player_effect = Estimate,
    lower = Q2.5,
    upper = Q97.5
  ) |>
  arrange(desc(player_effect))

player_effects |>
  slice_head(n = 15) |>
  gt()
```

```{r}
player_effects |>
  slice_max(player_effect, n = 20) |>
  mutate(player_name = fct_reorder(player_name, player_effect)) |>
  ggplot(aes(x = player_effect, y = player_name)) +
  geom_point() +
  geom_errorbarh(aes(xmin = lower, xmax = upper), height = 0.2) +
  labs(
    title = "Top adjusted player shooting effects",
    x = "Player random effect on log-odds scale",
    y = NULL
  )
```

# Demonstrating partial pooling

To show partial pooling, we compare raw field goal percentage with the Bayesian adjusted player effect. Players with fewer shots should generally be pulled more strongly toward the league average.

```{r}
partial_pooling_df <- player_raw |>
  inner_join(player_effects, by = "player_name")

ggplot(partial_pooling_df, aes(x = raw_fg_pct, y = player_effect, size = attempts)) +
  geom_point(alpha = 0.7) +
  labs(
    title = "Raw FG% versus Bayesian adjusted player effect",
    subtitle = "Point size represents number of attempts",
    x = "Raw field goal percentage",
    y = "Adjusted player effect on log-odds scale",
    size = "Attempts"
  )
```

# Posterior prediction

We now use the model to predict shot-making probabilities for specific hypothetical shot contexts.

```{r}
example_players <- player_effects |>
  slice_head(n = 3) |>
  pull(player_name)

new_shots <- crossing(
  player_name = factor(example_players, levels = levels(shots_model$player_name)),
  shot_distance = c(4, 12, 25)
) |>
  mutate(
    distance_z = (shot_distance - mean(shots_model$shot_distance)) / sd(shots_model$shot_distance),
    shot_type_simple = factor(
      if_else(shot_distance >= 23, "3PT", "2PT"),
      levels = levels(shots_model$shot_type_simple)
    ),
    is_home = factor("home", levels = levels(shots_model$is_home)),
    period = factor(1, levels = levels(shots_model$period)),
    clock_seconds_z = 0,
    score_margin_z = 0
  )
```

```{r}
preds <- fitted(
  fit_bayes,
  newdata = new_shots,
  re_formula = NULL,
  scale = "response",
  summary = TRUE
)

bind_cols(new_shots, as.data.frame(preds)) |>
  select(
    player_name,
    shot_distance,
    shot_type_simple,
    Estimate,
    Q2.5,
    Q97.5
  ) |>
  gt()
```

These predictions are posterior summaries of the probability of making a shot, accounting for uncertainty in both the fixed effects and player effects.

# Prediction intervals for future shot outcomes

Credible intervals describe uncertainty about the unknown probability of making a shot. Prediction intervals describe uncertainty in future observed outcomes. Because a single future field goal attempt is binary, the posterior predictive distribution is made of 0s and 1s. Therefore, single-shot prediction intervals can look wide and discrete even when the predicted make probability is estimated fairly precisely.

```{r}
set.seed(415)


y_pred_single <- posterior_predict(
  fit_bayes,
  newdata = new_shots,
  re_formula = NULL,
  ndraws = 4000
)

single_shot_prediction_intervals <- map_dfr(seq_len(ncol(y_pred_single)), function(j) {
  tibble(
    row_id = j,
    predicted_make_rate = mean(y_pred_single[, j]),
    pred_50_lower = quantile(y_pred_single[, j], 0.25),
    pred_50_upper = quantile(y_pred_single[, j], 0.75),
    pred_80_lower = quantile(y_pred_single[, j], 0.10),
    pred_80_upper = quantile(y_pred_single[, j], 0.90),
    pred_98_lower = quantile(y_pred_single[, j], 0.01),
    pred_98_upper = quantile(y_pred_single[, j], 0.99)
  )
})

new_shots |>
  mutate(row_id = row_number()) |>
  left_join(single_shot_prediction_intervals, by = "row_id") |>
  select(
    player_name,
    shot_distance,
    shot_type_simple,
    predicted_make_rate,
    pred_50_lower,
    pred_50_upper,
    pred_80_lower,
    pred_80_upper,
    pred_98_lower,
    pred_98_upper
  ) |>
  gt() |>
  tab_header(
    title = "Posterior prediction intervals for one future shot",
    subtitle = "Binary posterior predictive outcomes: 0 = miss, 1 = make"
  ) |>
  cols_label(
    predicted_make_rate = "Predicted Pr(make)",
    pred_50_lower = "50% lower",
    pred_50_upper = "50% upper",
    pred_80_lower = "80% lower",
    pred_80_upper = "80% upper",
    pred_98_lower = "98% lower",
    pred_98_upper = "98% upper"
  ) |>
  fmt_percent(columns = predicted_make_rate, decimals = 1)
```

For a single future shot, the prediction intervals answer: *What outcomes should we expect for one actual future attempt in this context?* Since the observed outcome can only be a make or a miss, these intervals often include both 0 and 1. That does not mean the model is uninformative; it reflects the fact that even high-probability shots can be missed and low-probability shots can be made. The predicted make rate is usually the most useful single-shot predictive summary.

To make prediction intervals easier to interpret, we also simulate the number of made shots out of 100 future attempts taken under the same context. This keeps the same model-based uncertainty but expresses prediction on a more stable basketball scale: made shots per 100 similar attempts.

```{r}
set.seed(415)


epred_draws <- posterior_epred(
  fit_bayes,
  newdata = new_shots,
  re_formula = NULL,
  ndraws = 4000
)

made_per_100_draws <- apply(
  epred_draws,
  2,
  function(p_draws) rbinom(length(p_draws), size = 100, prob = p_draws)
)

made_per_100_intervals <- map_dfr(seq_len(ncol(made_per_100_draws)), function(j) {
  tibble(
    row_id = j,
    predicted_makes_per_100 = mean(made_per_100_draws[, j]),
    pred_50_lower = quantile(made_per_100_draws[, j], 0.25),
    pred_50_upper = quantile(made_per_100_draws[, j], 0.75),
    pred_80_lower = quantile(made_per_100_draws[, j], 0.10),
    pred_80_upper = quantile(made_per_100_draws[, j], 0.90),
    pred_98_lower = quantile(made_per_100_draws[, j], 0.01),
    pred_98_upper = quantile(made_per_100_draws[, j], 0.99)
  )
})

new_shots |>
  mutate(row_id = row_number()) |>
  left_join(made_per_100_intervals, by = "row_id") |>
  select(
    player_name,
    shot_distance,
    shot_type_simple,
    predicted_makes_per_100,
    pred_50_lower,
    pred_50_upper,
    pred_80_lower,
    pred_80_upper,
    pred_98_lower,
    pred_98_upper
  ) |>
  gt() |>
  tab_header(
    title = "Posterior prediction intervals for makes out of 100 similar shots",
    subtitle = "50%, 80%, and 98% prediction intervals"
  ) |>
  cols_label(
    predicted_makes_per_100 = "Predicted makes per 100",
    pred_50_lower = "50% lower",
    pred_50_upper = "50% upper",
    pred_80_lower = "80% lower",
    pred_80_upper = "80% upper",
    pred_98_lower = "98% lower",
    pred_98_upper = "98% upper"
  ) |>
  fmt_number(
    columns = c(
      predicted_makes_per_100,
      pred_50_lower,
      pred_50_upper,
      pred_80_lower,
      pred_80_upper,
      pred_98_lower,
      pred_98_upper
    ),
    decimals = 1
  )
```

These intervals connect directly to the research question. The model is estimating how shot context and player-specific ability affect make probability, and the prediction intervals translate those estimated probabilities into expected future results. Shorter shots should generally have higher predicted makes per 100 attempts than longer shots, and players with larger positive random effects should have higher prediction intervals than comparable players taking the same shot. The 50% interval shows a typical range of future performance, the 80% interval gives a broader but still practically useful range, and the 98% interval shows the range of outcomes that would be plausible under nearly all posterior predictive draws.

# Posterior predictive checking

Posterior predictive checking compares simulated data from the fitted model to the observed data.

```{r}
pp_check(fit_bayes, ndraws = 100) +
  labs(title = "Posterior predictive check: overall distribution")
```

Because the response is binary, we also check whether the model reproduces group-level summaries such as make percentage by shot zone.

```{r}
yrep <- posterior_predict(fit_bayes, ndraws = 200)

obs_by_zone <- shots_model |>
  group_by(shot_zone_basic) |>
  summarize(obs_fg = mean(made), n = n(), .groups = "drop")

sim_by_zone <- map_dfr(seq_len(nrow(yrep)), function(s) {
  shots_model |>
    mutate(y_sim = yrep[s, ]) |>
    group_by(shot_zone_basic) |>
    summarize(sim_fg = mean(y_sim), .groups = "drop") |>
    mutate(draw = s)
})

zone_ppc <- sim_by_zone |>
  group_by(shot_zone_basic) |>
  summarize(
    sim_lower = quantile(sim_fg, 0.025),
    sim_median = median(sim_fg),
    sim_upper = quantile(sim_fg, 0.975),
    .groups = "drop"
  ) |>
  left_join(obs_by_zone, by = "shot_zone_basic")

ggplot(zone_ppc, aes(x = sim_median, y = fct_reorder(shot_zone_basic, sim_median))) +
  geom_point() +
  geom_errorbarh(aes(xmin = sim_lower, xmax = sim_upper), height = 0.2) +
  geom_point(aes(x = obs_fg), shape = 4, size = 3) +
  labs(
    title = "Posterior predictive check by shot zone",
    subtitle = "Dots are simulated medians; x-marks are observed field goal percentages",
    x = "Field goal percentage",
    y = "Shot zone"
  )
```

A good model should generate replicated zone-level field goal percentages that are close to the observed values. Large discrepancies suggest missing structure, such as omitted predictors or interactions.

# Model comparison

We compare the hierarchical model to a simpler non-hierarchical Bayesian logistic regression.

```{r}
fit_no_player <- brm(
  formula = model_formula_fixed,
  data = shots_model,
  family = bernoulli(link = "logit"),
  prior = priors_fixed_only,
  sample_prior = "yes",
  chains = 4,
  iter = 2500,
  warmup = 1000,
  seed = 415,
  control = list(adapt_delta = 0.97, max_treedepth = 12)
)
```

```{r}
loo_bayes <- loo(fit_bayes)
loo_no_player <- loo(fit_no_player)

loo_compare(loo_bayes, loo_no_player)
```

If the hierarchical model has better expected log predictive density, this supports the claim that player-level variation improves out-of-sample predictive performance.

# Optional richer model: player-specific distance effects

A more flexible hierarchical model allows the effect of distance to differ by player:

$$\text{logit}(p_i)
=
\alpha
+
u_{\text{player}[i]}
+
(\beta_1 + v_{\text{player}[i]})\text{distance}_{z,i}
+
\cdots.$$

This lets players vary not only in baseline shooting ability but also in how strongly distance affects their make probability.

```{r eval=FALSE}
fit_bayes_slope <- brm(
  made ~ distance_z + shot_type_simple + is_home + period +
    clock_seconds_z + score_margin_z + (1 + distance_z | player_name),
  data = shots_model,
  family = bernoulli(link = "logit"),
  prior = priors_main,
  sample_prior = "yes",
  chains = 4,
  iter = 4000,
  warmup = 1500,
  seed = 415,
  control = list(adapt_delta = 0.99, max_treedepth = 15)
)

loo_compare(loo(fit_bayes), loo(fit_bayes_slope))
```

We treat this as optional because it is computationally heavier and may require more data per player.

# Prior sensitivity analysis

A stronger Bayesian analysis should investigate whether conclusions depend strongly on the prior.

We compare the main weakly informative prior to a tighter prior and a wider prior.

```{r}
priors_tighter <- c(
  prior(normal(0, 1), class = "Intercept"),
  prior(normal(0, 0.5), class = "b"),
  prior(exponential(2), class = "sd")
)

priors_wider <- c(
  prior(normal(0, 3), class = "Intercept"),
  prior(normal(0, 2), class = "b"),
  prior(exponential(0.5), class = "sd")
)
```

```{r}
fit_tighter <- brm(
  formula = model_formula,
  data = shots_model,
  family = bernoulli(link = "logit"),
  prior = priors_tighter,
  chains = 4,
  iter = 2000,
  warmup = 1000,
  seed = 416,
  control = list(adapt_delta = 0.97, max_treedepth = 12)
)

fit_wider <- brm(
  formula = model_formula,
  data = shots_model,
  family = bernoulli(link = "logit"),
  prior = priors_wider,
  chains = 4,
  iter = 2000,
  warmup = 1000,
  seed = 417,
  control = list(adapt_delta = 0.97, max_treedepth = 12)
)
```

```{r}
sensitivity_fixed <- bind_rows(
  fixef(fit_bayes) |> as.data.frame() |> rownames_to_column("parameter") |> mutate(prior = "main"),
  fixef(fit_tighter) |> as.data.frame() |> rownames_to_column("parameter") |> mutate(prior = "tighter"),
  fixef(fit_wider) |> as.data.frame() |> rownames_to_column("parameter") |> mutate(prior = "wider")
)

sensitivity_fixed |>
  select(prior, parameter, Estimate, Q2.5, Q97.5) |>
  gt()
```

If the major conclusions, especially the signs and approximate magnitudes of the most important effects, are stable across these priors, then the analysis is not overly prior-driven.

# Frequentist comparison

For comparison, we fit a frequentist logistic mixed-effects model using `glmer()`.

```{r}
fit_freq <- glmer(
  made ~ distance_z + shot_type_simple + is_home + period +
    clock_seconds_z + score_margin_z + (1 | player_name),
  data = shots_model,
  family = binomial(link = "logit"),
  control = glmerControl(optimizer = "bobyqa")
)
```

```{r}
summary(fit_freq)
```

```{r}
freq_fixed <- tidy(fit_freq, effects = "fixed", conf.int = TRUE)

bayes_fixed <- fixef(fit_bayes) |>
  as.data.frame() |>
  rownames_to_column("term") |>
  mutate(term = if_else(term == "Intercept", "(Intercept)", term)) |>
  rename(
    bayes_estimate = Estimate,
    bayes_lower = Q2.5,
    bayes_upper = Q97.5
  )

comparison <- freq_fixed |>
  select(term, freq_estimate = estimate, freq_lower = conf.low, freq_upper = conf.high) |>
  left_join(bayes_fixed, by = "term")

comparison |>
  gt()
```

The frequentist model estimates similar log-odds coefficients, but the Bayesian model provides full posterior distributions and allows more direct probability statements. The Bayesian model also makes it natural to perform prior predictive checks and prior sensitivity analysis.

# Main results

The final results should be filled in after running the code. The main findings should discuss the following:

1.  **Shot distance:** Longer shots should generally have a lower posterior probability of being made.
2.  **Shot type:** Three-point attempts should generally be harder than two-point attempts after controlling for player differences.
3.  **Game context:** Home/away effects, score margin, and time remaining in the period may be smaller after adjusting for shot difficulty and player ability.
4.  **Player effects:** Some players have positive adjusted shooting effects, meaning they make shots at a higher rate than expected given shot context.
5.  **Partial pooling:** Low-attempt players are pulled closer to the league average than high-attempt players.
6.  **Model fit:** Posterior predictive checks should show whether the model captures overall and zone-level shot-making rates.
7.  **Prior sensitivity:** If results are similar under tighter and wider priors, conclusions are more robust.

# Discussion

This project demonstrates the value of Bayesian hierarchical modeling in basketball analytics. A raw field goal percentage ignores the difficulty of the shots a player takes. A player who attempts many difficult shots may appear inefficient using raw percentage alone, while another player who mostly attempts easier shots may appear more efficient than their true shot-making ability suggests.

Our hierarchical logistic regression adjusts for several components of shot difficulty while estimating player-specific ability. The hierarchical structure is important because it partially pools player estimates. Players with fewer attempts are shrunk toward the league average, while players with many attempts are estimated more independently from the data.

The model also gives posterior predictive distributions, allowing us to quantify uncertainty in predicted shot probabilities. This is more informative than simply reporting a fitted probability from a non-Bayesian model.

There are limitations. The available play-by-play data does not fully capture defender distance, contest level, pass quality, whether the shot was self-created, or a true shot-clock variable. In addition, our shot distance variable is partly derived from play text and coordinates rather than from official optical tracking data. These limitations may bias player effects if certain players consistently take more difficult shots that are not fully measured by the available variables.

# Conclusion

We used a Bayesian hierarchical logistic regression model to estimate the probability that an NBA shot is made. The model included shot context predictors and player-specific random effects. The Bayesian approach allowed us to quantify uncertainty, perform prior predictive checking, estimate partial pooling, generate posterior predictions, compare models, and examine sensitivity to prior assumptions.

The most important statistical contribution of the project is that it separates raw shot-making from adjusted shot-making. Rather than ranking players only by observed field goal percentage, the model estimates how much each player outperforms or underperforms expected shot probability after adjusting for shot context.

# References

- Bürkner, P.-C. (2017). `brms`: An R package for Bayesian multilevel models using Stan. *Journal of Statistical Software*.
- Stan Development Team. *Stan Modeling Language Users Guide and Reference Manual*.
- SportsDataverse. `hoopR`: Data and tools for basketball.
- ESPN NBA play-by-play data accessed through `hoopR`.
